A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-05 podc_paper/algorithm-2.tex
add algorithm2

\section{Queue Implementation} \label{DescriptQ}

\subsection{Overview}
We use a \emph{tournament tree} to agree on a total ordering of the operations performed on the queue.
The tree is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
assigned to each process. 
Each tree node  stores an array of \emph{blocks}, where each block represents a 
sequence of enqueues and a sequence of dequeues.
See Figure \ref{tournament} for an example.
In this section, we use an infinite array of blocks in each node.
Section \ref{reducing} describes how to replace the infinite array by a representation with bounded space.

To perform an operation on the queue, a process $P$ appends a new block containing that  
operation to the blocks array in $P$'s leaf.
Then, $P$ attempts to propagate the operation to each node along the path from that leaf to the root of the tree.
We shall define a total order on all operations that have been propagated to the root, which 
will serve as the linearization ordering of the operations.

To propagate an operation to a node $v$ in the tree, $P$ first observes
the blocks in both of $v$'s children that are not already in $v$,
creates a new block by combining information from those blocks, and attempts to append this 
new block to $v$'s blocks array using a CAS instruction.
Following \cite{}, we call this a 3-step sequence a
\op{Refresh} on $v$. %(see Figure \ref{fig::propagstep}).
A successful \op{Refresh} by $P$ may propagate pending operations by several other processes to $v$.
This cooperative approach is sufficient to ensure that if $P$ fails to CAS a 
new block into $v$'s array twice,
then its operation has been propagated to $v$ by some other process, so $P$ can continue 
onwards towards the root.

Now suppose $P$'s operation has been propagated all the way to the root.
If $P$'s operation is an enqueue, it has obtained a place in the linearization ordering and can terminate.
If $P$'s operation is a dequeue, $P$ must use information in the tree to compute the value that the
dequeue must return.  To do this, $P$ first determines which block in the root contains its operation
(since the operation may have been propagated to the root by some other process).
Then, $P$ determines whether the queue is empty when its dequeue is linearized. 
If so, it returns \nl\ and we call it a \emph{null dequeue}.
If not, $P$ computes the rank\footnote{We say that the $r$th element in a sequence has rank $r$ within that sequence.} $r$ of its dequeue in the linearization ordering
among all non-null dequeues,
finds the $r$th enqueue in the linearization, and returns that enqueue's value.

The primary challenge is thus figuring out what information to store in each block so that 
the following tasks can be done efficiently (in a polylogarithmic number of steps).
\begin{enumerate}[label={(T\arabic*)}]
\item
\label{construct}
Construct a block for node $v$ that represents the operations contained in $O(p)$ consecutive blocks in $v$'s children, as required for a \op{Refresh}.
\item
\label{findinroot}
Given a dequeue in a leaf that has been propagated to the root, find that operation's position in the root's blocks array.
\item
\label{findrank}
Given a dequeue's position in the root, determine whether it is a null dequeue (i.e., whether the queue is empty when it is linearized)
or determine the rank $r$ of the enqueue whose value it should return.
\item
\label{findenqueue}
Find the $r$th enqueue in the linearization ordering.
\end{enumerate}
Since these tasks depends on the linearization ordering, we describe that ordering next.

\begin{figure}[tb]
\input{tournamentTree.pdf_t}
\caption{An example tournament tree with four processes. 
We show explicitly the enqueue sequence and dequeue sequence represented by each block in the \fld{blocks} arrays of the seven nodes.  The leftmost element of each \fld{blocks} array is a dummy block.
Arrows represent the indices stored in \eleft\ and \eright\ fields of blocks (as described in Section \ref{sec:fields}).
The fourth process's Deq\sub{6} is still in progress.
The linearization order for this tree is
Enq(a) Enq(e) Deq\sub{2} $\mid$ Enq(b) Deq\sub{4} Deq\sub{5} $\mid$ Enq(d) Enq(f) Enq(h) Deq\sub{1} $\mid$ Enq(c) Deq\sub{3} $\mid$ Enq(g), where vertical bars indicate boundaries of blocks in the root.\label{tournament}}
\end{figure}

\begin{figure}
\input{implicit.pdf_t}
\caption{\label{implicit}The actual, implicit representation of the tree shown in Figure \ref{tournament}.
The leaf blocks simply show the \fld{element} field.
Internal blocks show the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields,
and \eleft\ and \eright\ fields are shown using arrows as before.
Root blocks also have the additional \fld{size} field.
The \fld{super} field is not shown.}
\end{figure}

\subsection{Linearization Ordering}

We shall show that an operation terminates only after a block containing it has been appended to the root's blocks array.
So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
$op_1$ will be in an earlier block than $op_2$ in root's blocks array.
Thus, our linearization orders operations according to the block they belong to in the root's blocks array.
Operations that appear in the same block are necessarily concurrent, so we can choose how to order them.

Each block in a leaf represents a single operation.
Each block $B$ in an internal node $v$ results from merging
several consecutive blocks from each of $v$'s children.
The merged blocks in $v$'s children are called the \emph{direct subblocks} of $B$.
A block $B'$ in a descendant of $v$ is a \emph{subblock} of $B$ if it is a direct subblock of $B$
or a subblock of a direct subblock of $B$.
A block $B$ represents the set of operations in all of $B$'s subblocks in leaves of the tree.

The operations propagated by a \op{Refresh} are all pending when the \op{Refresh} occurs,
so there is at most one operation per process.
Hence, a block represents at most $p$ operations in total.  
Moreover, we never append empty blocks, so 
each block represents at least one operation and it follows that a block can have at most $p$ direct subblocks.

As mentioned above, we are free to order operations within a block however we like.
For convenience, we order the enqueues and dequeues separately, and put the 
operations that were propagated from the left child before the operations from the right child.
More formally, we inductively define an sequences $E(B)$ and $D(B)$ of the enqueues and dequeues
represented by a block $B$.
If $B$ is a block in a leaf representing an enqueue operation, its enqueue sequence $E(B)$ is that operation
and its dequeue sequence $D(B)$ is empty.  If $B$ is a block in a leaf representing a dequeue, $D(B)$ is that single operation and $E(B)$ is empty.
If $B$ is a block in an internal node $v$ with direct subblocks $B^L_1, \ldots, B^L_\ell$ from the left child of $v$
and $B^R_1,\ldots,B^R_r$ from the right child of $v$, then $B$'s operation sequences are defined by the concatenations 
\begin{eqnarray}
E(B) &=& E(B^L_1)\cdots E(B^L_\ell)\cdot E(B^R_1) \cdots E(B^R_r)\nonumber\\
D(B) &=& D(B^L_1)\cdots D(B^L_\ell)\cdot D(B^R_1) \cdots D(B^R_r)\label{defSeqs}
\end{eqnarray}
We say the block $B$ \emph{contains} the operations in $E(B)$ and $D(B)$.

Finally, when designing a total order for the operations propagated to the root, we choose
to put each block's enqueues before its dequeues.
Thus, if the root's blocks array contains blocks $B_1, \ldots, B_k$, the 
linearization ordering is 
\begin{equation}
L=E(B_1)\cdot D(B_1) \cdot E(B_2) \cdot D(B_2) \cdots E(B_k) \cdot D(B_k).
\label{linearization}
\end{equation}

\subsection{Designing a Block Representation to Solve Tasks \ref{construct} to \ref{findenqueue}}
\label{sec:fields}

Each \node\ of the tournament tree has an infinite array called \fld{blocks}.
To simplify the code, each \node\ has an empty \block\ $B_0$ in \fld{blocks}[0].
We define $E(B_0)$ and $D(B_0)$ to be the empty sequence.
Each \node\ also has a \fld{head} index that stores the position in the \fld{blocks} array to be used
for the next attempt to append a \block\ to this \node.

If a block contained an explicit representation of its sequences of enqueues and dequeues,
it would take $\Omega(p)$ time to construct a block, which would be too slow for task \ref{construct}.
Instead, the block stores an implicit representation of the sequences.
We now explain how we designed the fields for this implicit representation. 
Refer to Figure \ref{implicit} for an example showing how the tree in Figure \ref{tournament} is actually represented, and Figure \ref{object-fields} for the definitions of the fields of \blocks\ and \nodes.

The information stored in a \block\ depends on whether it is in an internal node or a leaf.
A block in a leaf represents a single enqueue or dequeue.  The \fld{element} field stores the value
enqueued for an enqueue operation, or \nl\ if the operation is a dequeue.

\here{If space permits, we might want to add some examples in the following paragraphs that refer back to Figure \ref{implicit}.}

Each block in an internal \node\ $v$ stores the indices of the last direct subblock in $v$'s left and right child in the fields \eleft\ and \eright.  
Thus, the direct subblocks of $v.\fld{blocks}[b]$ are
\begin{eqnarray}\label{defsubblock}
v.\fld{left.blocks}[v.\fld{blocks}[b-1].\fld{end\sub{left}}+1..v.\fld{blocks}[b].\fld{end\sub{left}}] \mbox{ and}\nonumber\\
v.\fld{right.blocks}[v.\fld{blocks}[b-1].\fld{end\sub{right}}+1..v.\fld{blocks}[b].\fld{end\sub{right}}].
\end{eqnarray}
This allows us to navigate to the direct subblocks of any block easily.
Blocks also store prefix sums of the numbers of enqueues and dequeues:
the block in $v.\fld{blocks}[b]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
that store the total number of enqueues and dequeues in $\var{v}.\fld{blocks}[1..\var{b}]$.
These fields allow us to pinpoint the exact location of an operation among the subblocks of a given block.
For example, consider finding the $r$th enqueue in the linearization ordering.
We use the \fld{sum\sub{enq}} fields of the root's blocks to do a binary search
to locate which block in the root contains the operation.
If we know which block $B$ in some \node\ $v$ contains the enqueue,
we can use the \fld{sum\sub{enq}} field again to determine which child of $v$ contains the enqueue
and then to do a binary search
among the direct subblocks of $B$ in that child.
Thus, we work our way down the tree until we find a leaf block, which explicitly stores 
the enqueue.
%Now, suppose we want to find the $r$th enqueue in the linearization ordering for task \ref{findenqueue}.Let $B_1, B_2, \ldots, B_k$ be the blocks in the root.
%First, we need prefix sums of the number of enqueues in $E(B_1)\cdot E(B_2)\cdots E(B_i)$
%so that we can do a binary search for the block $B_e$ that contains the $r$th enqueue.
%This prefix sum also allows us to know the rank $r'$ within $E(B_e)$ of the $r$th enqueue.
%Once we have $r'$, we need the number of enqueues that $B_e$ received from its left child
%to determine whether the enqueue came from the left or right child of the root.
%Suppose the enqueue came from the right child $v_r$.
%Then, we know that the index of the block in $v_r$ that contains the enqueue
%is between $B_{e-1}.\eright + 1$ and $B_e$.\eright.
%We can again do a binary search within this range.
%For this, we can again use the prefix sums of the number of enqueues in any prefix of the array $v_r.blocks$.
%We can then continue in this way down the tree until reaching a leaf where the enqueue is stored explicitly.
We shall show that the binary search in the root can be done in $O(\log p + \log q)$ steps,
and the binary search within each other \node\ along the path to a leaf can be done in $O(\log p)$ steps,
for a total of $O(\log^2 p + \log q)$ steps to complete task \ref{findenqueue}.
All of the information needed for this search process can be derived from the 
\eleft, \eright\ and \fld{sum\sub{enq}} fields.

A block is called the \emph{superblock} of all of its direct subblocks.
To facilitate task \ref{findinroot}, each block $B$ has a field \fld{super} that contains
the (approximate) index of its superblock in the parent \node's \fld{blocks} array.
We shall ensure that the value of $B.\fld{super}$ differs from the true index of $B$'s superblock by at most 1.
This allows a process to determine the true location of the superblock by checking the \eleft\ or \eright\ values of just two \blocks\ in the parent \node.
Thus, starting from an operation in a leaf's block, one can use these indices to track the 
operation all the way up the path to the root, and determine the operation's location in a root block
in $O(\log p)$ time.

We now consider task \ref{findrank}.
First, we must determine whether the queue is empty when a dequeue occurs.
To facilitate this, each block in the root stores a \fld{size} field that stores the number of elements
in the queue after all operations in the linearization ordering up to that block (inclusive) 
have been performed.
We can easily determine which dequeues in a block $B_d$ in the root are null dequeues using
$B_{d-1}.\fld{size}$, which stores the size of the queue just before $B_d$'s operations are performed, and the number of enqueues and dequeues in $B_d$.
We can also compute the number of non-null dequeues in blocks $B_1, \ldots, B_{d-1}$ 
as $B_{d-1}.\fld{sum\sub{enq}}-B_{d-1}.\fld{size}$.
Then, for any non-null dequeue in $B_d$, we can use this information to determine its
rank among all non-null dequeues in the linearization ordering, and hence the rank of the enqueue
whose value it should return (among all enqueues).

Having defined the fields required for tasks \ref{findinroot}, \ref{findrank} and \ref{findenqueue},
we can easily see how to construct a new block $B$ during a \op{Refresh} in $O(1)$ time.
A \op{Refresh} on \node\ $v$ reads the values $h_{\ell}$ and $h_{r}$ of the \fld{head} fields of $v$'s children and stores 
$h_{\ell}-1$ and $h_{r}-1$ in $B.\eleft$ and $B.\eright$.
Then, we can compute $B.\fld{sum\sub{enq}}$ as $v.\fld{left}.\fld{blocks}[B.\eleft].\fld{sum\sub{enq}} + v.\fld{right}.\fld{blocks}[B.\eright].\fld{sum\sub{enq}}$.
For a block $B$ in the root, $B.\fld{size}$ can be computed using the \fld{size} field of the previous block $B'$ and
the number of enqueues and dequeues in $B$:
$B.\fld{size} = \max(0, B'.\fld{size} + (B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}}) - (B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}}))$.

The only remaining field is $B.\fld{super}$.  When the block 
$B$ is created for a \node\ $v$, we do not yet know where its
superblock will eventually be installed in $B$'s parent.
So, we leave $B.\fld{super}$ blank.  After $B$ is installed 
in $v.\fld{blocks}[\var{h}]$, processes cooperate to fill in $B.\fld{super}$ 
when they attempt to advance $v.\fld{head}$ from $h$ to $h+1$,
using a value they read from the \fld{head} field of $v$'s parent.
As mentioned above, this might not be the exact index of $B$'s superblock, but we
shall prove that it is close.

\renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENTSINGLE}{\com\ #1}}
\begin{figure}
\here{Does \var{leaf} have to be shared for helping when doing GC?}
\begin{algorithmic}[1]
\setcounter{ALG@line}{1}
\Statex \linecomment{Shared variable}
\begin{itemize}
\item \typ{Node} \var{root} \Comment{root of a binary tree of \tt{Node}s with one leaf for each process}
\end{itemize}

@@

\Function{void}{Enqueue}{\typ{Object} \var{e}} 
    \State \hangbox{let \var{n} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
		\fld{sum\sub{enq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}+1,\\
		\fld{sum\sub{deq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}}\label{enqNew}
    \State \Call{Append}{\var{n}}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue()}{} 
    \State \hangbox{let \var{n} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
	    \fld{sum\sub{enq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}},\\
	    \fld{sum\sub{deq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1}\label{deqNew}
    \State \Call{Append}{\var{n}}
    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, 1}\label{invokeIndexDequeue}
    \State \Return{ \Call{FindResponse}{\var{b, i}}}\label{deqRest}
\EndFunction{Dequeue}

@@

\Function{void}{Append}{\typ{Block} \var{B}} 
    \State \linecomment{append block to leaf and propagate to root}
    \State \var{leaf.}\fld{blocks}[\var{leaf.}\head] \assign\ \var{B}\label{appendLeaf}
    \State \var{leaf.}\head\ \assign\ \var{leaf.}\head+1 \label{appendEnd} 
    \State \Call{Propagate}{\var{leaf.}\fld{parent}} 
\EndFunction{Append}

@@

\Function{void}{Advance}{\typ{Node} \var{v}, \typ{int} \var{h}} 
    \State \linecomment{set \var{v}.\fld{blocks}[\var{h}].\fld{super} and }
    \State \linecomment{increment \var{v}.\fld{head} from \var{h} to \var{h}+1}
    \If{$\var{v}\neq \var{root}$}
	    \State \var{h\sub{p}} \assign\ \var{v}.\fld{parent}.\head \label{readParentHead}
    	\State \Call{CAS}{\var{v}.\fld{blocks}[\var{h}].\fld{super}, \nl, \var{h\sub{p}}} \label{setSuper1}
	\EndIf
    \State \Call{CAS}{\var{v}.\head, \var{h}, \var{h}+1} \label{incrementHead}
\EndFunction{Advance}

@@

\begin{algorithmic}[1]
\setcounter{ALG@line}{50}

\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
    \State\linecomment{create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
    \State let \var{new} be a new \typ{Block} \label{initNewBlock}
    \State \var{new}.\eleft \assign\ $\var{v}.\fld{left}.\head - 1$\label{createEndLeft}
    \State \var{new}.\eright \assign\ $\var{v}.\fld{right}.\head - 1$\label{createEndRight}
	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}\label{createSumEnq}
	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}\label{createSumDeq}
    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnq}
    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
    \If{$\var{v} = \var{root}$}
        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLength}
    \EndIf
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}\label{testEmpty}
        \State \Return \nl \hfill \com\ no blocks need to be propagated to \var{v}
    \Else
        \State \Return \var{new}
    \EndIf
\EndFunction{CreateBlock}

@@

\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{return $\langle\var{b}', \var{i}'\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$}
    \State \linecomment{is $(\var{i}')$th dequeue of $D(\var{root}.\fld{blocks}[\var{b}'])$}
    \State \linecomment{Precondition: $\var{v}.\fld{blocks}[\var{b}]$ is not \nl, has been }
    \State \linecomment{propagated to root and contains at least \var{i} dequeues}
    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCase}
    \Else
	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
    	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$} \var{sup} \assign\ $\var{sup}+1$\label{supertest}
	    \EndIf
	    \State \linecomment{compute index \var{i} of dequeue in superblock}
	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ \\
	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}\label{computeISuperStart}
        \If{$\fld{dir} = \fld{right}$} 
        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
					$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRight}
        \EndIf \label{computeISuperEnd}
        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, \var{sup}, \var{i}}
    \EndIf
\EndFunction{IndexDequeue}

@@

\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    %  $i\geq 1$ and \var{root}.\fld{blocks}[\var{b}] is non-null and}
    %  \State \linecomment{contains at least \var{i} dequeues}
    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}\label{FRNum}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmpty}
        \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNull}
    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeE}
		\State \linecomment{compute enqueue's block using binary search}
		\State find min $b_e\leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRb}
		\State \linecomment{find rank of enqueue within its block}
		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$\label{FRi}
        \State \Return \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswer}
    \EndIf
\EndFunction{FindResponse}

@@

\begin{figure}
\begin{algorithmic}[1]
\setcounter{ALG@line}{102}

\Function{element}{GetEnqueue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \Comment{returns argument of \var{i}th enqueue in $E(\var{v}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Preconditions: $\var{i}\geq 1$ and \var{v}.\fld{blocks}[\var{b}] is non-\nl\ and contains at least \var{i} enqueues}

>>>>

\section{Garbage Collection}

\renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENTSINGLE}{\com\ #1}}
\begin{figure}
%\here{Does \var{leaf} have to be shared for helping when doing GC?}
\begin{algorithmic}[1]
\setcounter{ALG@line}{1}
\Statex \linecomment{Shared variable}
\begin{itemize}
\item \typ{Node} \var{root} \Comment{root of a binary tree of \tt{Node}s with one leaf for each process} % TODO: indent these lines in order to be more readable. I tries some times and i could not figure out doing that in algorithm.
\item \gc{\typ{*Node[]} \var{leafOfProcess[1..p]} \Comment{Pointer to the leaf assigned to each process.}}
\item \gc{\typ{int[]} \var{lastDequeuedByProcess[1..p]}}
\end{itemize}

\vspace{0.8em} %TODO: clean code

@@

\Function{void}{Enqueue}{\typ{Object} \var{e}} 
    \State \hangbox{let \var{n} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
		\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}}+1,\\
		\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}}}
    \State \Call{Append}{\var{n}}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue()}{} 
    \State \hangbox{let \var{n} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
	    \fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}},\\
	    \fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}}+1}
    \State \Call{Append}{\var{n}}
    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, \var{leaf}.\head, 1}
    \State \Return{ \Call{FindResponse}{\var{b, i}}}
    \label{deqRest}
\EndFunction{Dequeue}

@@

\Function{void}{Append}{\typ{Block} \var{B}} 
    \State \linecomment{append block to leaf and propagate to root}
    \State \var{leaf}.\fld{blocks}[\var{leaf}.\head] \assign\ \var{B}\label{appendLeaf}
    \gc{\State \var{leaf}.\fld{blocks}.\Call{TryAppend}{\var{B}, \var{leaf}.\head}\label{appendLeafGC}}
    \State \var{leaf}.\head\ \assign\ \var{leaf}.\head+1 \label{appendEnd} 
    \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
\EndFunction{Append}

@@

\Function{void}{Advance}{\typ{Node} \var{v}, \typ{int} \var{h}} 
    \State \linecomment{set \var{v}.\fld{blocks}[\var{h}].\fld{super} and }
    \State \linecomment{increment \var{v}.\fld{head} from \var{h} to \var{h}+1}
    \State \var{h\sub{p}} \assign\ \var{v}.\fld{parent}.\head \label{readParentHead}
    \State \Call{CAS}{\var{v}.\fld{blocks}[\var{h}].\fld{super}, \nl, \var{h\sub{p}}} \label{setSuper1}
    \State \Call{CAS}{\var{v}.\head, \var{h}, \var{h}+1} \label{incrementHead}
\EndFunction{Advance}

@@

\begin{algorithmic}[1]
\setcounter{ALG@line}{48}

\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
    \State\linecomment{create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
    \State let \var{new} be a new \typ{Block} \label{initNewBlock}
    \State \var{new}.\eleft \assign\ $\var{v}.\fld{left}.\head - 1$\label{createEndLeft}
    \State \var{new}.\eright \assign\ $\var{v}.\fld{right}.\head - 1$
	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}
	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnq}
    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
    \If{$\var{v} = \var{root}$}
        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLength}
    \EndIf
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}
        \State \Return \nl \hfill \com\ no blocks need to be propagated to \var{v}
    \Else
        \State \Return \var{new}
    \EndIf
\EndFunction{CreateBlock}

@@

\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{return $\langle\var{x, y}\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$}
    \State \linecomment{is \var{y}th dequeue of $D(\var{root}.\fld{blocks}[\var{x}])$}
    \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] has been propagated to root}
    \State \linecomment{and $D(\var{v}.\fld{blocks}[\var{b}])$ has at least \var{i} dequeues}
    \If{$\var{v} = \var{root}$}
        \State\Return $\langle\var{b, i}\rangle$ \label{indexBaseCase}
    \Else
	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
    	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$}
	        \State{\var{sup} \assign\ $\var{sup}+1$}
	    \EndIf
	    \State \linecomment{compute index \var{i} of dequeue in superblock}
	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ \\
	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}
        \If{$\fld{dir} = \fld{right}$} 
        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
					$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRight}
        \EndIf \label{computeISuperEnd}
        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, \var{sup}, \var{i}}
    \EndIf
\EndFunction{IndexDequeue}

@@

\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{find response to \var{i}th dequeue in \var{root}.\fld{blocks}[\var{b}]}
    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmpty}
    \gc{\State \var{lastDequeuedByProcess[this process id]} \assign\ \var{b}}
        \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNull}
            \EndIf
\State \linecomment{response is the \var{e}th enqueue in the root}
        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeE}
		\State \linecomment{compute enqueue's block using binary search}
		\State find min $\var{b}'\leq \var{b}$ with $\var{root}.\fld{blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{e}$
		\State \linecomment{find rank of enqueue within its block}
		\State $\var{i}' \assign\ \var{e} - \var{root}.\fld{blocks}[\var{b}'-1].\fld{sum\sub{enq}}$
        \State \var{e} \assign\ \Call{GetEnqueue}{\var{root}, $\var{b}'$, $\var{i}'$}\label{findAnswer}
        \gc{\State \var{lastDequeuedByProcess[this process id]} \assign\ $\var{b}'$}
        \State \Return{\var{e}}

\EndFunction{FindResponse}

@@

\begin{figure}
\begin{algorithmic}[1]
\setcounter{ALG@line}{101}

\Function{element}{GetEnqueue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \Comment{returns argument of \var{i}th enqueue in $E(\var{v}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Precondition: $\var{i}\geq 1$ and $E(\var{v}.\fld{blocks}[\var{b}])$ contains at least \var{i} enqueues}
